Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-05 E-Hypergraphs-LICS24.tex
Update E-Hypergraphs-LICS24.tex

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}

The correspondence of the previous section further extends to include the equations of a monoidal theory: the equations of a theory are considered as rewrite rules on string diagrams,  which are then modelled as DPO-rewrites on hypergraphs with interfaces. 
DPO rewriting is a standard technique for formalising graph rewriting categorically. 
In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $L \rightsquigarrow R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $L$ within $\mathcal{G}$,  remove it,  and then plug $R$ into the hole.  In DPO rewriting,  such a rewrite rule is formalized as a span $L \xleftarrow{} K \xrightarrow{} R$ in some category $\mathcal{C}$ with pushouts,  where $K$ is some invariant that should be maintained during rewriting.
CHRIS: "WITHIN G",  BUT WE ARE IN AN ARBITRARY CATEGORY: NOT NECESSARILY GRAPHS?

A typical DPO square looks as follows,  where the two marked squares are pushouts. 
\[
\begin{tikzcd}
L \arrow[d, "m"'] & K \arrow[d] \arrow[l, "f"'] \arrow[r, "g"] & R \arrow[d]   \\
G^{\urcorner}     & C \arrow[l] \arrow[r]                      & ^{\ulcorner}H
\end{tikzcd}
\]
Morphisms $K \to C \to G$ are called a \textit{pushout complement} and morphism $m$ is called a \textit{match},  defined when the corresponding pushout complement exists. When $f,g$ are expected to be monomorphisms, constructing $C$ can be intuitively understood as finding an occurrence of $L$ in $G$ and removing from this occurrence everything that has no pre-image in $K$ and then inserting in there everything from $R$ that has no pre-image in $K$ -- hence the two pushouts: one for deletion and one for insertion.  In this case,  we say that $G$ is rewritten into $H$ via the rewrite rule $L \xleftarrow{} K \xrightarrow{} R$.

To express the rewriting of (hyper-)graphs with interfaces, e.g. the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout with \textit{interfaces} rewriting (DPOI).

\begin{definition}[DPOI rewriting]
Fix a category $\mathbb{C}$ with pushouts.
Given morphisms $G \xleftarrow{} J$ and $H \xleftarrow{} J$ in $\mathbb{C}$, $G$ rewrites into $H$ with interface $J$ if there exists a rewrite rule $L \xleftarrow{} K \xrightarrow{} R$ and object $C$ with suitable morphisms in $\mathbb{C}$ such that the diagram below commutes and two marked squares are pushouts.
\[
\begin{tikzcd}
L \arrow[d, "m"'] & K \arrow[d] \arrow[l, "f"'] \arrow[r, "g"] & R \arrow[d]   \\
G^{\urcorner}     & C \arrow[l] \arrow[r]                      & ^{\ulcorner}H \\
                  & J \arrow[lu] \arrow[u] \arrow[ru]          &              
\end{tikzcd}
\]

\end{definition}
When $\mathbb{C} = \catname{Hyp_{\Sigma}}$, the rewrite rules become cospans of hypergraphs with matching interfaces: $i \xrightarrow{} L \xleftarrow{} j$, $i \xrightarrow{} R \xleftarrow{} j$.
By noting that a cospan $0 \xrightarrow{} L \xleftarrow{[f_{i},f_{j}]} i + j$ encodes the same data as a cospan $i \xrightarrow{f_{i}} L \xleftarrow{f_{j}} j$ (Remark 3.14~\cite{Frobenius}) we can formulate a rewrite rule as a single span $L \xleftarrow{} K \xrightarrow{} R$ where $K \cong i + j$.
Likewise, for the target cospan $n \xrightarrow{} G \xleftarrow{} m$,  $J \cong n + m$.
% However, this notion of DPO does not account for hypergraphs with interfaces, i.e. those from $\MdaCospans$ where $l,r$ from $\langle l, r \rangle$ are cospans. To support this we first assume that we only need to maintain the interfaces of $i \xrightarrow{} L \xleftarrow{} j$ and $i \xrightarrow{} R \xleftarrow{} j$ during rewriting, then we can take $K \cong i + j$, i.e. take $K$ to be the coproduct of $i,j$ (remark 3.14~\cite{Frobenius}). What remains is to account for the interfaces in $G$, $C$, and $H$, which brings us to the following modified DPOI square. $J \cong n + m$ where $n$ and $m$ are the input and output interfaces of $G$, $C$, and $H$ respectively. The only difference from DPO square above is that we now explicitly require that the initial object $G$, intermediate object $C$ and final object $H$ all share the same interfaces.

CHRIS: DEFINE DPOI FORMALLY,  IN DEFINITION ENVIRONMENT? PERHAPS COMBINE WITH BOUNDARY COMPLEMENT BELOW,  TWO FIGURES ON ONE LINE?

To mimic rewrites in $\textbf{PROP}(\Sigma)$ with the help of DPOI rewriting in $\catname{Hyp_{\Sigma}}$, we need to put some restrictions both on pushout complement and on matching as not all of them correspond to valid rewrites in $\textbf{PROP}(\Sigma)$. CHRIS: "LEAD TO"? BE PRECISE! :)

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For monogamous cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} G \xleftarrow{b_2} m$
and mono $f : L \to G$, a pushout complement as depicted below

@@

\begin{definition}
Given $G \xleftarrow{} n+m$ and $H \xleftarrow{} n + m$ in $\catname{Hyp_{\Sigma}}$, $G$ rewrites convexly into $H$ with interface $n + m$ --- notation $(G \xleftarrow{} n + m ) \Rrightarrow_{\mathcal{R}}  (H \xleftarrow{} n + m )$ --- if there exist rule $L \xleftarrow{} i + j \xrightarrow{} R$ in the set of rewrite rules $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} C \xleftarrow{} n+m$ such that the DPOI diagram above commutes and its marked squares are pushouts and the following conditions hold

\begin{itemize}
    \item $m : L \to G$ is a convex match;
    \item $i + j \to C \to G$ is a boundary complement in the leftmost pushout.
\end{itemize}

By resorting to convex matching and boundary complement we virtually require that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.

ALEKSEI: THE WORD `CONVEXLY' ABOVE MEANS THAT THE REWRITING IS FORMULATED FOR MOMOGAMOUS MORPHISMS. SHALL WE STICK TO THIS FORMULATION?

\section{E-Hypergraphs}

>>>>

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
The correspondence of the previous section means that the symmetric monoidal equations of $\textbf{SMT}(\Sigma, \mathcal{E})$ are absorbed in the hypergraph representation.  The equations $\mathcal{E}$ are then to be considered as rewrite rules on string diagrams,  which are then modelled as DPO-rewrites on hypergraphs with interfaces. 
DPO rewriting is a standard technique for formalising graph rewriting categorically. 
In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalized as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square looks as follows,  where the two marked squares are pushouts. 
\[
\begin{tikzcd}
\mathcal L \arrow[d, "m"'] & \mathcal K \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
\mathcal G^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner} \mathcal H
\end{tikzcd}
\]

The morphism $\mathcal L \to \mathcal G$ is called a \textit{match},  defined whenever there is a corresponding \textit{pushout complement} -- the pair of morphisms $\mathcal K \to \mathcal C \to \mathcal G$ completing the top half of the pushout square defined by $\mathcal K \xrightarrow{} \mathcal L \xrightarrow{} \mathcal G$. When the morphisms involved in the rewrite rule are monomorphisms,  as is usually required,  constructing $\mathcal C$ can be intuitively understood as finding an occurrence of $\mathcal L$ in $\mathcal G$ and removing from this occurrence everything that has no pre-image in $\mathcal K$.  The final graph $\mathcal H$ is the recovered by inserting into the hole everything from $\mathcal R$ that has no pre-image in $\mathcal K$ -- hence the two pushouts: one for deletion and one for insertion.  In this case,  we say that $\mathcal G$ is rewritten into $\mathcal H$ via the rewrite rule $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$.

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{Frobenius}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  Thus we reach the following variant of DPO rewriting. 
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
\[
\begin{tikzcd}
\mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
\mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
                  & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
\end{tikzcd}
\]
\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acylcicity.

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For monogamous cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement as depicted in the square below

@@

\begin{definition}[Convex DPOI rewriting]
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} \mathcal C \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $m : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal C \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.

CHRIS: WANT TO STATE THE CORRESPONDENCE THEOREM WITH SMTS HERE? NEED TO DEFINE $\lceil-\rceil$ AND INTERPRETATION? Hmm
\begin{theorem}[Theorem 35~\cite{Frobenius2}]
For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
\[
	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
\]
where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
\end{theorem}
\section{E-Hypergraphs}
